\documentclass[12pt]{article}
\usepackage{a4wide}
\newcommand{\GAP}{\textsf{GAP}}
\newcommand{\bd}{\begin{description}}
\newcommand{\ed}{\end{description}}
\parindent=0pt
\parskip=\medskipamount
\title{Extending GASMAN to non-contiguous workspaces}
\author{Steve Linton}
\begin{document}
\maketitle
\section{Introduction}

These are some notes on an idea that may or may not get implemented.

Persistent problems arise in \GAP\ because GASMAN needs a contiguous
block of memory for the workspace, which may need to grow
unpredictably as a calculation continues. Standard memory allocation
schemes using \verb|malloc| or \verb|sbrk| cannot readily support
this.  The typical problem is that an initial workspace is allocated
with \verb|sbrk|, then some library function calls \verb|malloc|
causing \verb|malloc| to do an \verb|sbrk| to enlarge its arena, then
GASMAN tries to enlarge the workspace and is given a block of memory
beyond the \verb|malloc| arena, which is not a contiguous extension of 
the existing workspace.

The workspace cannot be moved as a whole (using \verb|realloc| or \verb|mremap|)
because bag identifiers are pointers to the masterpointers of the 
bags they identify. This could be changed, but might slow down all bag
accesses.

There are therefore two possibilities:
\begin{enumerate}
\item Remove the requirement for a non-contiguous workspace \label{sacks}
\item Implement our own versions of \verb|malloc| (etc.) which
supplies memory from within the workspace. \label{ourmalloc}
\end{enumerate}

Blocks allocated by the \verb|malloc| replacement under option
\ref{ourmalloc} cannot be moved, and so become rigid obstacles within
the workspace. This basically takes us into option \ref{sacks}.

We therefore consider to follow option \ref{sacks}.

\section{Basic Design}

We aim to modify GASMAN without going to the trouble of a complete
rewrite, so that it can cope with a workspace that consists of a
number of separate memory extents, which we call \emph{sacks}. There
is a frequent requirement in GASMAN to recognise whether an arbitrary
value is a valid bag id. Part of the check involves conparisons with
the upper and lower bounds of the masterpointer area. To avoid slowing 
this check down, we insist that the masterpointer area be entirely
within one sack (the \emph{master sack}).

There will be at most one sack capable of being extended
contiguously, the last one allocated. We call this the \emph{top sack}. 

Each sack will be divided into up to seven parts (in order):
\bd
\item[A Sack header] of fixed size, containing information about the sack
\item[Masterpointers] Only in the master sack
\item[Reserved Allocatable Space] space which will not be used for new 
bags immediately, but may become Freely Allocatable after a partial
garbage collection
\item[Freely Allocatable Space] space in which new bags may be created
\item[New Bags] bags created since the last garbage collection
\item[Old Bags] bags which have survived a garbage collection
\item[Unallocatable Space] space which will not be used for new 
bags immediately, but may become Reserved Allocatable Space after the
next full garbage collection
\ed

The change of order from the current workspace is to allow the size of 
the masterpointer area to vary without necessarily having to move all
the bags.

All sacks will be linked in a circular doubly-linked list.  Sacks
which contain (more that 4KB of) allocatable space are called
\emph{open} (otherwise \emph{closed}). All open sacks are linked in a
second circular doubly linked list. The sack in which the first
attempt at the next allocation will be made is called the
\emph{current allocation sack}.




\section{Algorithms}

\subsection{Initialisation}

The initial workspace should be a single sack, which will be the
master sack. 

We could explore allocating it very large, but refraining 
from accessing most of it. Under Linux, at least, this would work
rather well, and might make this whole scheme unnecessary.

\subsection{Allocating a bag}

To allocate a new bag, we proceed:
\begin{enumerate}
\item If there is enough space in the current allocation sack then
allocate there
\item Otherwise go around the linked list of open sacks and see if any 
of them have room
\item If so, allocate there and make that sack the ``current
allocation sack''
\item If not garbage collect
\end{enumerate}

\subsection{Garbage collection}

The mark phase of garbage collection is unchanged.

The current sweep phase runs destination and source pointers through
the Young bags area consolidating the live bags into what will become the
Old bags area. In a full garbage collection, all bags are considered
young. After that, the workspace and masterpointer area sizes
are adjusted, which may involve moving the entire bags area. 

We could sweep each sack separately. This is simple, but fragments the 
free space, and we will need to tackle moving bags between sacks
anyway, to cater for adjustments in the size of the masterpointer
area.

I therefore propose that the sweeping process should run through all
the (non-trivial) Young Bags areas with the aim of consolidating the
surviving bags into as few sacks as (quickly) possible, and, if
possible, of filling those sacks full enough that they can be regarded 
as closed. The master sack should be the first choice to be open, when 
possible, in case it becomes necessary to expand the masterpointer area.

This becomes tricky when we get close to the end of a sack into which
we are sweeping, and can only fit in smaller bags. If the source
pointer points at too large a bag, however, we can look ``ahead'' of
the source pointer for such bags, copy them, and remark them as dead
so that we don't sweep them up again. Then, when the destination
pointer moves into the next sack, we can resume sweeping normally from
the source pointer.

The next important phase of garbage collection is the adjustment of
the sizes of things. The following considerations arise:

\begin{itemize}
\item We must provide a single block of free space of the size
requested when GC was invoked. If we cannot do this, we must fail.
\item We must provide a free masterpointer if one is
needed. Otherwise, we must fail.
\item We want to provide a ``reasonable'' amount of freely allocatable space, 
so that we will not have to garbage collect again too soon.
\item After a full garbage collection, we may want to provide a
``reasonable'' amount of reserved allocatable space  to avoid having
to do a full GC again too soon.
\item We may wish to return unused space to the OS. On most systems
this is only important if the space is dirty (ie we have written to
it).
\item We wish to ensure that the amount of space reserved for
masterpointers is appropriate. Since this space must be in the master
sack, we may have to move other bags out of that sack to do this, or
we may get some free space in a sack where we might not want it.
\end{itemize}

The calculations for this phase can be done before the sweep phase,
but until we have swept, there may not be space to move bags around.

\section{Parameters}

I propose that we set all our parameters in multiples of 4KB. This is
accurate enough for all sensible purposes, keeps everything inside 28
bits, allowing the parameters to be manipulated in \GAP, and will
encourage things to lie on page boundaries, which will suit most OSs.

\bd
\item[InitialSackSize] The total size at which the master sack should
start. If this memory cannot be allocated, the initialization
fails. Default 8MB.
\item[InitialUsedSize] The amount of the master sack which should
\emph{not} be marked unallocatable to start with. Default is
InitialSackSize, values greater than InitialSackSize are reduced with
warning.
\item[MinAllocationWindow] If this much allocatable space cannot be found
after a partial garbage collection then d a full garbage
collection. Default 512KB
\item[MaxAllocationWindow] If this is non-zero, then garbage collection
will try to ensure that the freely allocatable space is in a single
region of this size. If this is zero, then no 
reserved allocatable space will ever be created. Default zero. If
non-zero, should not be less than MinAllocationWindow.
\item[AbsoluteSizeMaximum] A strict top limit on the total size of all
sacks. Default 128MB. A limit of zero means no limit.
\item[MinMasterPointerRatio] The space for free masterpointers should
be kept at least this fraction (expressed in 1000ths) of the total
allocatable space. Default 166
\item[MaxMasterPointerRatio] The space for free masterpointers should
be kept at most this fraction (expressed in 1000ths) of the total
allocatable space. Default 250. A value less than or equal to
MinMasterPointerRatio is an error.
\item[MinAllocatableSpaceRatio] The total allocatable space should be
kept to at least this fraction of the total currently
allocated. Default 125.
\item[MinAllocatableSpace] The total allocatable space should also be
kept to at least this absolute amount. Default 0
(MinAllocatableSpaceRatio is used instead)
\item[MaxAllocatableSpaceRatio] If the total allocatable space is
larger than this ratio of the total allocated space in all sacks and
there is no unallocatable space in the top sack, and the top sack has
allocatable space in it, then some of that space may be returned to
the OS. Default 250
\item[MinimumSizeChangeRatio] When the top sack is being expanded or
shrunk, always change it by at least this proportion of the total size 
of all sacks. Default 100
\item[MiniumuSizeChange]  When the top sack is being expanded or
shrunk, always change it by at least this absolute amount. Default
512KB
\item[MinimumSackSize] The minimum size with which any new top sack will be
allocated, or to which it will be allowed to shrink. Default 1MB.
\item[MinimumSackRatio] The minimum fraction of the total size of all
sacks with which any new top sack will be
allocated, or to which it will be allowed to shrink. Default 250.
\item[MaximumIgnoredSpace] The maximum allocatable space that may be
left in a sack without the sack being regarded as open. Default 4KB.
\ed


\section{Size Change Procedures}

\subsection{After a Partial Garbage Collection}

Let $x$ be the amount of space required from this garbage collection

If no sack contains amount $x$ of allocatable space then do a full
garbage collection.

If AllocationWindow is non-zero then 
\begin{enumerate}
\item If any sack contains $x + \mbox{MaxAllocationWindow}$ 
of allocatable space, then 
mark that as freely allocatable and the rest as reserved
allocatable. If there is more than one such bag, avoid choosing the
master bag.
\item If not, Let y be the total amount of allocatable space in all sacks,
except those containing less that MaximumIgnoredSpace
\item If $y > x + \mbox{MaxAllocationWindow}$, then mark $x +
\mbox{MaxAllocationWindow}$ of space
freely allocatable, and the remaining allocatable space reserved
allocatable. If there is choice, avoid the master bag.
\item If not, and $y > \mbox{MinAllocationWindow}+x$ then mark all allocatable
space freely allocatable
\item If not then do a full collection
\end{enumerate}
otherwise:
\begin{enumerate}
\item  Let $y$ be the total amount of allocatable space in all sacks,
except those containing less that MaximumIgnoredSpace
\item If $y >  \mbox{MinAllocationWindow}+x$ then mark all allocatable space 
 freely allocatable
\item If not then do a full collection
\end{enumerate}

Now check that the number of free masterpointers exceeds
$y*(\mbox{MinMasterPointerRatio}/(1000* \mbox{word length}))$. 
If not, do a full garbage collection.


Set up the linked list of open sacks and the Current Allocation Sack. 

Return.

\subsection{After a Full Garbage Collection}

Let $x$ be the amount of space required. 

Let $y$ be the total amount of allocatable space remaining after the
garbage collection, discounting those sacks containing less than
MaxIgnoredSpace.

Let $z$ be the number of free masterpointers after the garbage
collection. Let $m$ be the size of a masterpointer.

Let $w$ be the total space in all allocated bags.

Let $n$ be the number of allocated bags, so that $mn$ is the number of 
bytes of masterpointers in use.

Let $t$ be the total size of all sacks.


If no sack has as much as $x$ allocatable space in it then we will 
need to make special arrangements (the large bag case). We consider
this at the end.


If \begin{eqnarray}
m(z-1) &\ge& \mbox{MinMasterPointerRatio} * (y-x) / 1000\label{minmp}\\
m(z-1) &\le& \mbox{MaxMasterPointerRatio} * (y-x) / 1000\label{maxmp}\\
y -x&\ge& \mbox{MinAllocatableSpace}\label{minspace}\\
y -x&\ge& \mbox{MinAllocatableSpaceRatio}*w/1000\label{minspacer}\\
y -x&\le& \mbox{MaxAllocatableSpaceRatio}*w/1000\label{maxspacer}
\end{eqnarray}

then the current sizes are fine.

We next seek a solution to the inequalities simply by changing the
size of the masterpointer area. 

We consider increasing the number of masterpointers by $a$ (which may
be negative). 

We find that
\begin{eqnarray}
ma &\ge& {\mbox{MinMasterPointerRation} * (y-x) \over 1000 +
\mbox{MinMasterPointerRation}} - m(z-1) \\
ma &\le& {\mbox{MaxMasterPointerRation} * (y-x) \over 1000 +
\mbox{MaxMasterPointerRation}} - m(z-1) \\
ma & \le & y - x - \mbox{MinAllocatableSpace}\\
ma & \le & y - x - \mbox{MinAllocatableSpaceRatio} * w /1000\\
ma & \ge & y - x - \mbox{MaxAllocatableSpaceRatio} * w /1000
\end{eqnarray}

Let $y_s$ be the amount of allocatable space in 
the master sack. 
There is an additional constraint, whose exact form depends on
whether any sack except the master sack has $x$ bytes free. If so
then it is simply:

\begin{equation}
ma \le y_s
\end{equation}

otherwise it is:

\begin{equation}
ma \le y_s -x
\end{equation}

If these inequalities has a range of solutions then take the center
point. It then simply remains to change the pointers in the master
sack, and, perhaps, initialize the new masterpointers and sort out the 
division between freely allocatable and reserved allocatable. The sweep phase 
was organised so that the free space should be concentrated in the
master sack, so we are unlikely to miss a solution which might involve 
moving bags out of the master sack.

Otherwise, we have to consider changing the size of the whole
workspace, by making some unallocatable space allocatable, extending
the top sack, shrinking the top sack, or allocating a new sack.


We therefore consider increasing the workspace by $b$ bytes (which
might be negative) and get a series of simultaneous inequalities in
$a$ and $b$, which had better always have a solution.


If \begin{eqnarray}
ma &\ge& \mbox{MinMasterPointerRatio} * (y+b-x-ma) / 1000- m(z-1)\\
ma &\le& \mbox{MaxMasterPointerRatio} * (y+b-x-ma) / 1000 - m(z-1)\\
b-ma&\ge& \mbox{MinAllocatableSpace} + x-y\\
b -ma&\ge& \mbox{MinAllocatableSpaceRatio}*w/1000 + x- y\\
b -ma&\le& \mbox{MaxAllocatableSpaceRatio}*w/1000 + x -y
\end{eqnarray}


We can solve this by taking a middle value for $b-ma$ from the
solution range of the last three, and then using that to fix a range
for $a$, using the first two, taking the center value from that range
and then solving for $b$. There is an additional constraint for $a$
if the master sack is not the top sack, namely that the master
pointers all have to fit into the master sack.

Now we adjust $b$ (towards $+\infty$) to avoid the open ranges
$(\mbox{MinimumSizeChange},0)$, $(0,\mbox{MinimumSizeChange})$
$(\mbox{MinimumSizeChangeRatio}*t/1000,0)$ and
$(0,\mbox{MinimumSizeChangeRatio}*t/1000)$.

We could go back and re-solve for $a$, but this is probably not worth
doing, the next GC will catch it.

If $b$ is negative, and the top sack contains no unallocatable space,
then we can try to reduce the size of the top sack, subject to leaving 
at least MinimumSackSize and at least
$t*\mbox{MinimumSackRation}/1000$. This may require moving some
bags out of the top sack. If this proves difficult (due to large bags, 
presumably) then we simply revise $b$ upwards (possibly to 0).


Otherwise, we look at the top sack and see if it contains unallocatable
space. If it does, and contains at least $b$ bytes, then we can make
$b$ bytes of unallocatable space allocatable, and we are done. If it
contains less than $b$, but more than 0, then we make it all
allocatable, and reduce $b$ subject to avoiding the four intervals above.. 

Next we try to extend the top sack by $b$ bytes. If we succeed then we 
are fine.

If not, we need to allocate a new sack. We may need to increase $b$,
so that it is not less than MinimumSackSize and not less than
$t*\mbox{MinimumSackRatio}/1000$. If no existing sack has total size
as large as $x$ bytes (special rules in the master sack), then we must 
also ensure that $b > x$.

Having got the extra space, we now may have to move bags between sacks 
for two reasons:
\begin{itemize}
\item To make room for extra masterpointers
\item To make room for the new bag of size $x$ which started this
problem
\end{itemize}

If the workspace is very clogged with large bags in relatively small
sacks, then we may find ourselves in trouble. In this (hopefully rare)
situation, we expand the top sack, or allocate a new sack, big enough
to take all the bags from the master sack and the new sack to be
allocated (or possibly a bit less than that).  If we can't manage to
do this then we give up and suggest that the user restart with a
larger master sack.



If we fail to allocate the space we want, then we should simply take
what we can get, and see if we can fit in the new bag, and divide up
any remaining space betwee free space and free masterpointers.

\end{document}