\Chapter{Group Libraries}

When you start {\GAP}, it  already knows several groups. Currently {\GAP}
initially knows the following groups:
\beginlist
  \item{$\bullet$} some basic groups, such  as cyclic groups or symmetric
    groups (see "Basic groups"),
  \item{$\bullet$}  a library of  primitive  permutation groups of degree
    $\<1000$, complete  for degrees  $\<256$ (see "Primitive  permutation
    groups"),
  \item{$\bullet$}  the  irreducible solvable subgroups of  $GL(n,p)$ for
    $n>1$ and $p^n \< 256$ (also in "Primitive permutation groups"),
  \item{$\bullet$} the transitive permutation groups of degree at most 23
    (see "Transitive permutation groups"),
  \item{$\bullet$} the small groups of size at most 1000, excluding sizes
    512 and 768 (see "Small groups"),
  \item{$\bullet$}  the  finite perfect  groups of   size  at most $10^6$
    (excluding 11 sizes) (see "Finite perfect groups").
%\item{$\bullet$} the irreducible maximal finite integral matrix groups 
%  of dimension at most  24  (see  "Irreducible Maximal Finite Integral
%  Matrix Groups"),
%\item{$\bullet$} the  crystallographic  groups  of  dimension  at most  4 
%  (see "The Crystallographic Groups Library").
\endlist

There   is usually no   relation  between the   groups  in the  different
libraries. Note that a system administrator may choose to install all, or
only a  few,  or even  none of the  libraries.  So some  of the libraries
mentioned below may not be available on your installation.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Section{Basic groups}\null

\>CyclicGroup( \[ Is\dots Group \[ and \dots\ \] \], <n> )
\)CyclicGroup( IsMatrixGroup, <field>, <n> )

`CyclicGroup( <n> )' returns a cyclic group of size <n>. If you specify a
property (or conjunction  of properties) as  optional first argument, the
resulting group will  have  these properties. For  example, `CyclicGroup(
IsPcGroup,  <n> )'  will return a   group described by  a pc presentation
(with as  many  generators as  <n>  has prime  divisors). This method  of
giving  an additional   property  works  for  all  construction functions
described in  this section. If you  specify `IsMatrixGroup', you can also
prescribe a field   as additional second  argument,   otherwise a default
field will  be  chosen    (which is    the  rationals  in  the case    of
`CyclicGroup'). If you  leave out the property  completely, `CyclicGroup'
assumes `IsPermGroup'.
\beginexample
    gap> c12 := CyclicGroup( 12 );
    Group( ( 1, 2, 3, 4, 5, 6, 7, 8, 9,10,11,12) )
    gap> c12 := CyclicGroup( IsMatrixGroup, 3 );
    Group( [ [ [ 0, 1, 0 ], [ 0, 0, 1 ], [ 1, 0, 0 ] ] ], ... )
    gap> c12 := CyclicGroup( IsMatrixGroup, GF(4), 3 );
    Group( 
    [ [ [ 0*Z(2), Z(2)^0, 0*Z(2) ], [ 0*Z(2), 0*Z(2), Z(2)^0 ], [ Z(2)^0, 
              0*Z(2), 0*Z(2) ] ] ], ... )
    gap> c105 := CyclicGroup( IsPcGroup, 5*3*7 );
    Group( c105_1, c105_2, c105_3 )
\endexample
The following functions  also return permutation groups, unless  directed
otherwise.

\>AbelianGroup( \[ Is\dots Group \[ and \dots\ \] \], <sizes> )

returns an  abelian group  isomorphic  to $C_{sizes[1]} *  C_{sizes[2]} *
\ldots  *   C_{sizes[n]}$,  where <sizes>  must   be a  list  of positive
integers.

\>DihedralGroup( \[ Is\dots Group \[ and \dots\ \] \], <n> )
\>ElementaryAbelianGroup( \[ Is\dots Group \[ and \dots\ \] \], <n> )

return the corresponding group of size <n>.

\>PolyhedralGroup( \[ Is\dots Group \[ and \dots\ \] \], <p>, <q> )

returns the polyhedral group of size $<p> * <q>$. The numbers <p> and <q>
must be positive  integers and there must exist  a nontrivial <p>-th root
of unity modulo every prime factor of <q>.

\>SymmetricGroup( \[ Is\dots Group \[ and \dots\ \] \], <d> )
\>AlternatingGroup( \[ Is\dots Group \[ and \dots\ \] \], <d> )

return the symmetric resp.\ alternating group of degree <d>.

\>GeneralLinearGroup( \[ Is\dots Group \[ and \dots\ \] \], <n>, <q> )
\>SpecialLinearGroup( \[ Is\dots Group \[ and \dots\ \] \], <n>, <q> )

These  functions return matrix groups by  default,  namely the general or
special linear group in its natural representation.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Section{Selection functions}

\)\fmark All<Library>Groups( $<fun>_1$, $<val>_1$, \dots\ )

For each  group library there  is a  *selection function*.  This function
allows you to select all groups from the library that have a given set of
properties.

The name of the selection  functions always begins  with `All' and always
ends with `Groups'. Inbetween is a  name that hints at  the nature of the
group library. For example, the selection function for the library of all
primitive  groups of degree    at  most 50 (see  "Primitive   permutation
groups") is called `AllPrimitiveGroups'.

These  functions take an  arbitrary number of  pairs of  arguments.   The
first argument in such a pair  is a function  that can  be applied to the
groups in the library,  and the second  argument is either a single value
that this  function must return in  order  to have this group included in
the selection, or a list of such values. For example,
\beginexample
    AllPrimitiveGroups( DegreeOperation,  [10..15],
                        Size,             [1..100],
                        IsAbelian,        false    );
\endexample
should return a list of all primitive  groups with  degree between 10 and
15 and size less than 100 that are not abelian.

Thus  the `AllPrimitiveGroups'  behaves  as if   it was implemented  by a
function similar to the one defined below, where `PrimitiveGroupsList' is
a list of all primitive groups.  Note, in  the definition below we assume
for simplicity that `AllPrimitiveGroups' accepts exactly 4 arguments.  It
is of course  obvious how to change this  definition so that the function
would accept a variable number of arguments.
\begintt
    AllPrimitiveGroups := function( fun1, val1, fun2, val2 )
        local    groups, g, i;

        groups := [];
        for i  in [ 1 .. Length( PrimitiveGroupsList ) ] do
            g := PrimitiveGroupsList[i];
            if      fun1(g) = val1  or IsList(val1) and fun1(g) in val1
                and fun2(g) = val2  or IsList(val2) and fun2(g) in val2
            then
                Add( groups, g );
            fi;
        od;
        return groups;
    end;
\endtt
Note that the real  selection functions are considerably  more difficult,
to improve the efficiency. Most  important, each recognizes a certain set
of functions and handles those properties using an index.

\)\fmark One<Library>Group( $<fun>_1$, $<val>_1$, \dots\ )

If you  do not want  all groups with the  prescribed properties, but only
one example, you should call the *example function*. It returns `fail' if
no such group exists in the library.

The following sections describing a particular group library mention only
the *extraction   function*, by which  you  address a group   via (1)~its
degree  or   size    (or for  matrix   groups  its    dimension  and  its
characteristic) and  (2)~its  number  in  the list  of all   groups  with
property~(1).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Section{Primitive permutation groups}

{\GAP} contains a library of primitive permutation groups which includes,
up to  isomorphism of permutation groups   (i.e., up to conjugacy  in the
corresponding symmetric group)
\beginlist
  \item{$\bullet$} the  non-affine primitive permutation groups of degree
    $\<1000$,   described    in  \cite{DixonMortimer88},  with generators
    calculated in \cite{Theissen97},
  \item{$\bullet$} all  primitive  permutation groups of  degree $\<256$,
    in particular,
    \itemitem{$\circ$} the primitive permutation groups up to degree~50,
      calculated by C.~Sims,
    \itemitem{$\circ$}  the solvable (hence affine) primitive permutation
      groups of degree $\<256$, calculated by M.~Short \cite{Short92},
    \itemitem{$\circ$} the insolvable affine primitive permutation groups
      of degree $\<256$, calculated in \cite{Theissen97}.
\endlist
Note that the affine primitive permutation groups of degrees 256--999 are
not included.

\>PrimitiveGroup( <deg>, <nr> )
\>IrreducibleSolvableGroup( <dim>, <p>, <nr> )
\>PrimitiveAffinePermGroupByMatrixGroup( <M> )
\>NrPrimitiveGroups( <deg> )
\>NrAffinePrimitiveGroups( <deg> )
\>NrSolvableAffinePrimitiveGroups( <deg> )

`PrimitiveGroup' returns the primitive permutation  group of degree <deg>
with number <nr>  from the list. If  $<deg>\ge 256$, there  are no affine
primitive  groups in  the list, so   you have to  skip  the rest  of this
paragraph. Otherwise, first  come the one-dimensional affine  groups over
the field $F_p$, i.e., Frobenius groups of the  form $ F_p\mathbin:A$ for
a  subgroup $A\le{\rm Aut}(F_p)$. (Short did  not include these groups in
his list.) Then come the other solvable  affine groups, in the same order
as in Short's list in {\sf GAP~3}. They  are constructed with the command
`PrimitiveAffinePermGroupByMatrixGroup(      <M>  )'   from   a  solvable
irreducible matrix group <M>, which in turn is obtained by the well-known
command `IrreducibleSolvableGroup'. Next  in the list come the insolvable
affine primitive  permutation groups   which are  analogously constructed
from insolvable irreducible matrix groups.

Now come the   non-affine primitive permutation  groups  of degree <deg>.
They have been  classified  into cohorts in  \cite{DixonMortimer88},  and
{\GAP}    represents a     cohort   as a     homomorphism   $\kappa\colon
N=N_{S_{<deg>}}(S)\to A$ whose kernel $S$  is the socle  of $N$ and every
primitive group in that cohort is the preimage of a subgroup of $A$ (only
one from   each conjugacy  class)  under $\kappa$.   For the  degrees  in
question,  $A$ is solvable. All  primitive groups in  the cohort $\kappa$
have the same socle, namely~$S$. The groups  of each cohort appear in the
list consecutively.

If  you want to  find out where  the non-affine  part  (or the insolvable
affine part) of a list begins, you  can use the number functions `Nr\dots
PrimitiveGroups'.

\)AllPrimitiveGroups( IsPrimitiveAffine, true, IsSolvable, true, %
                      $<fun>_3$, $<val>_3$, \dots\ )

If you use  the selection functions  for the primitive groups library and
select  only solvable affine  groups (as in  the example above), then you
can (in  the   place of the dots)    also give functions  that  you would
normally give   to `AllIrreducibleSolvableGroups'.  These functions  will
then be applied to the  point stabilizer of the affine  group, which is a
matrix group  from  M.~Short's library.  Hence you will  get only  groups
$V.M$  where  $V$ is  a    vector space and    $M$  a group  returned  by
`AllIrreducibleSolvableGroups( $<fun>_3$, $<val>_3$, \dots\ )'.

\>PrimitiveGroupSims( <deg>, <nr> )
\>AllPrimitiveGroupsSims( $<fun>_1$, $<val>_1$, \dots\ )
\>OnePrimitiveGroupSims( $<fun>_1$, $<val>_1$, \dots\ )

For  compatibility with earlier versions  of {\GAP}, the original list of
Sims, with the same numbers and the names given by Buekenhout and Leemans
\cite{BuekenhoutLeemans96},  is also   included.  It is accessed  by  the
function  `PrimitiveGroupSims'. You can also  use the selection functions
with this old list, their names end with `Sims'.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Section{Transitive permutation groups}

The transitive groups library contains representatives for all transitive
permutation groups of degree at most  23.  Two permutations groups of the
same degree are considered to be equivalent, if there is a renumbering of
points, which maps one group into the other one.  In other words, if they
lie in the save  conjugacy class  under operation  of the  full symmetric
group by conjugation.

\>TransitiveGroup( <deg>, <nr> )

`TransitiveGroup'  returns the <nr>-th transitive  group of degree <deg>.
Both  <deg> and <nr> must be  positive integers. The transitive groups of
equal  degree are  sorted with  respect to   their  size, so for  example
`TransitiveGroup(  <deg>, 1 )' is a  transitive group  of degree and size
<deg>, e.g, the cyclic  group  of size <deg>,   if <deg> is a  prime. The
arrangement of  the groups, the  generators and their names correspond to
the lists in \cite{ConwayHulpkeMcKay97}

`AllTransitiveGroups'  and `OneTransitiveGroup'   recognize the following
functions and get the corresponding properties from a precomputed list to
speed up  processing:

`NrMovedPoints', `Size',   `Transitivity', and `IsPrimitive'.  You do not
need to pass  those functions first, as  the selection function picks the
these properties first.

This library was computed by Gregory Butler, John McKay, Gordon Royle and
Alexander  Hulpke. The list  of transitive  groups  up to  degree  11 was
published in \cite{ButlerMcKay83}, the list of degree 12 was published in
\cite{Royle87}, degree 14 and   15 were published in \cite{Butler93}  and
degrees 16--22 in \cite{Hulpke96}.

The library was  brought into {\GAP} format by  Alexander  Hulpke, who is
responsible for all mistakes.
\beginexample
    gap> TransitiveGroup(10,22);
    S(5)[x]2
    gap> l:=AllTransitiveGroups(NrMovedPoints,12,Size,1440,IsSolvable,false);
    [ S(6)[x]2, M_10.2(12) = A_6.E_4(12) = [S_6[1/720]{M_10}S_6]2 ]
    gap> List(l,IsSolvable);
    [ false, false ]
\endexample

\>TransitiveIdentification( <G> )

Let <G> be a permutation group, acting transitively on a set  of up to 23
points.  Then `TransitiveIdentification' will return the position of this
group in the transitive  groups library.  This means,  if <G> operates on
$m$ points and    `TransitiveIdentification'  returns $n$,  then <G>   is
permutation isomorphic to the group `TransitiveGroup(m,n)'.
\beginexample
    gap> TransitiveIdentification(Group((1,2),(1,2,3)));
    2
\endexample

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Section{Finite perfect groups}
\index{perfect groups}%

The {\GAP} library of finite  perfect groups provides, up to isomorphism,
a list of all perfect groups whose sizes are less than  $10^6$  excluding
the following sizes:
\beginlist
  \item{$\bullet$}
      For $n = 61440$, 122880, 172032, 245760, 344064, 491520, 688128, or
      983040,  the perfect groups  of size  $n$  have not completely been
      determined yet.  The library  neither provides  the number of these
      groups nor the groups themselves.
  \item{$\bullet$}
      For  $n = 86016$,  368640,  or  737280,  the library  does not  yet
      contain  the perfect groups  of size  $n$,  it  only provides their
      their numbers which are 52, 46, or 54, respectively.
\endlist

Except for these eleven sizes, the list of altogether 1096 perfect groups
in the library  is complete. It relies   on results of  Derek~F. Holt and
Wilhelm  Plesken which are published  in their book  {\it Perfect Groups}
\cite{HoltPlesken89}. Moreover,   they have supplied  us  with files with
presentations of 488 of the groups. In terms  of these, the remaining 607
nontrivial groups in the library can be described as 276 direct products,
107 central  products, and 224  subdirect   products. They  are  computed
automatically by suitable {\GAP} functions whenever they are needed.

We are grateful to Derek Holt and Wilhelm Plesken for making their groups
available to the {\GAP} community  by contributing their files. It should
be noted that  their book contains a  lot of further information for many
of the library groups.  So we would like  to recommend  it to any  {\GAP}
user who is interested in the groups.

The library has been brought into {\GAP} format by Volkmar Felsch.

\>PerfectGroup( <selector> \[, <size> \[, <n> \] \] )

`PerfectGroup' is  the essential extraction function  of  the library. It
returns   a group, $G$   say, which is  isomorphic   to the library group
specified by   the  size number  [<size>,<n>]  or   by the  two  separate
arguments $size$ and  $n$,  assuming a default   value  of $n =  1$.  The
<selector> defines the representation,  in  which the group is  returned.
Possible selectors so far are `IsPermGroup' and `IsSubgroupFpGroup'
\beginexample
    gap> G := PerfectGroup(IsPermGroup,6048,1);  
    U3(3)
\endexample
The  generators   and  relators used    coincide   with  those  given  in
\cite{HoltPlesken89}.

As  all groups are stored  by presentations, a permutation representation
is obtained by coset enumeration. Note that some of the library groups do
not have a   faithful   permutation  representation  of small     degree.
Computations in these groups may be rather time consuming.
\beginexample
    gap> G:=PerfectGroup(IsPermGroup,129024,2);
    L2(8) N ( 2^6 E 2^1 A ) C 2^1
    gap> NrMovedPoints(G);
    14336
\endexample

\>PerfectIdentification(<G>)

Besides the  standard attributes for   a perfect group in the  respective
representation, each  groups  obtained  from   the library has   set  the
attribute `PerfectIdentification' to $[<size>,<nr>]$.

\>NumberPerfectGroups( <size> )

`NumberPerfectGroups' returns the number of non-isomorphic perfect groups
of size $size$ for  each positive integer  $size$ up to $10^6$ except for
the eight  sizes listed at the beginning  of  this section for  which the
number is not yet known. For these values as well as for any argument out
of range it returns the value $-1$.

\>NumberPerfectLibraryGroups( <size> )

`NumberPerfectLibraryGroups' returns the number of perfect groups of size
$size$ which are available in the  library of finite perfect groups. (The
purpose  of the function is  to provide a simple way  to formulate a loop
over all library groups of a given size.)

\>SizeNumbersPerfectGroups( <factor1>, <factor2>, \dots\ )

`SizeNumbersPerfectGroups' returns  a list of the  size and  index of all
library groups that contain the specified factors among their composition
factors. Each argument must either  be the name of a  simple group or  an
integer expression  which is the   product of the sizes   of one or  more
cyclic factors. (In fact, the function replaces any list of more than one
integer expression among the arguments by their product.)

The following text strings are accepted as simple group names.
\beginlist
  \item{$\bullet$} `A<n>' or  `A(<n>)' for the  alternating groups $A_n$,
    $5\leq n\leq9$, for example `A5' or `A(6)'.
  \item{$\bullet$}  `L<n>(<q>)'   or  `L(<n>,<q>)' for  $PSL(n,q)$, where
    $n\in\{2,3\}$ and $q$ a prime power, ranging
    \itemitem{$\circ$} for $n=2$ from 4 to 125
    \itemitem{$\circ$} for $n=3$ from 2 to 5
    \itemitem{$\circ$} for $n=2$ from 4 to 125
  \item{$\bullet$} `U<n>(<q>)'  or  `U(<n>,<q>)' for   $PSU(n,q)$,  where
    $n\in\{3,4\}$ and $q$ a prime power, ranging
    \itemitem{$\circ$} for $n=3$ from 3 to 5
    \itemitem{$\circ$} for $n=4$ from 2 to 2
  \item{$\bullet$} `Sp4(4)' or `S(4,4)' for the symplectic group $S(4,4)$,
  \item{$\bullet$} `Sz(8)' for the Suzuki group $Sz(8)$,
  \item{$\bullet$} `M<n>'  or `M(<n>)' for  the  Mathieu groups $M_{11}$,
    $M_{12}$, and $M_{22}$, and
  \item{$\bullet$} `J<n>' or `J(<n>)'   for  the Janko groups  $J_1$  and
    $J_2$.
\endlist

Note that,   for  most of  the  groups, the   preceding  list offers  two
different names in order  to  be consistent with   the notation  used  in
\cite{HoltPlesken89} as  well   as   with the   notation  used  in    the
`DisplayCompositionSeries' command of {\GAP}.  However, as the  names are
compared  as text strings, you are  restricted  to the above choice. Even
expressions like `L2(32)' or `L2(2\^5)' are not accepted.

As the use of the  term $PSU(n,q)$ is  not  unique in the literature,  we
state that here  it denotes the factor  group of $SU(n,q)$ by its centre,
where $SU(n,q)$ is  the group of all $n  \times n$ unitary matrices  with
entries in $GF(q^2)$ and determinant 1.

The purpose  of the function is  to provide a  simple way to  formulate a
loop over all library groups which contain certain composition factors.

\>DisplayInformationPerfectGroups( \[ <size> \[, <n> \] \] )

`DisplayInformationPerfectGroups' displays  some  information about   the
library group   $G$,  say,  which  is  specified    by the   size  number
[<size>,<n>] or by the two  arguments $size$ and $n$.   If, in the second
case, <n> is omitted,  the function will  loop over all library groups of
size <size>.

The information provided for $G$ includes the following items:
\beginlist
  \item{$\bullet$} a headline  containing the size number  $[<size>,<n>]$
    of $G$ in the form $<size>.<n>$ (the suffix $.<n>$ will be suppressed
    if, up to isomorphism, $G$ is the only perfect group of size $size$),
  \item{$\bullet$} a message if $G$ is simple  or quasisimple, id est, if
    the factor group of $G$ by its centre is simple,
  \item{$\bullet$} the ``description'' of  the structure of  $G$ as it is
    given by Holt and Plesken in \cite{HoltPlesken89} (see below),
  \item{$\bullet$} the size of  the centre of $G$  (suppressed, if $G$ is
    simple),
  \item{$\bullet$} the prime decomposition of the size of $G$,
  \item{$\bullet$} orbit sizes for  a faithful permutation representation
    of $G$ which is provided by the library (see below),
  \item{$\bullet$} a reference to each occurrence of $G$ in the tables of
    section 5.3    of  \cite{HoltPlesken89}. Each  of   these  references
    consists of a class number and an internal number $(i,j)$ under which
    $G$ is listed in that class. For some groups, there  is more than one
    reference because these groups belong to more than one of the classes
    in the book.
\endlist
\beginexample
gap> DisplayInformationPerfectGroups( 30720, 3 );
|#I Perfect group 30720:  A5 ( 2^4 E N 2^1 E 2^4 ) A
|#I   size = 2^11*3*5  orbit size = 240
|#I   Holt-Plesken class 1 (9,3)
|_
gap> DisplayInformationPerfectGroups( 30720, 6 );
|#I Perfect group 30720:  A5 ( 2^4 x 2^4 ) C N 2^1
|#I   centre = 2  size = 2^11*3*5  orbit size = 384
|#I   Holt-Plesken class 1 (9,6)
|_
gap> DisplayInformationPerfectGroups( Factorial( 8 ) / 2 );
|#I Perfect group 20160.1:  A5 x L3(2) 2^1
|#I   centre = 2  size = 2^6*3^2*5*7  orbit sizes = 5 + 16
|#I   Holt-Plesken class 31 (1,1) (occurs also in class 32)
|#I Perfect group 20160.2:  A5 2^1 x L3(2)
|#I   centre = 2  size = 2^6*3^2*5*7  orbit sizes = 7 + 24
|#I   Holt-Plesken class 31 (1,2) (occurs also in class 32)
|#I Perfect group 20160.3:  ( A5 x L3(2) ) 2^1
|#I   centre = 2  size = 2^6*3^2*5*7  orbit size = 192
|#I   Holt-Plesken class 31 (1,3)
|#I Perfect group 20160.4:  simple group  A8
|#I   size = 2^6*3^2*5*7  orbit size = 8
|#I   Holt-Plesken class 26 (0,1)
|#I Perfect group 20160.5:  simple group  L3(4)
|#I   size = 2^6*3^2*5*7  orbit size = 21
|#I   Holt-Plesken class 27 (0,1)
\endexample

For any library  group  $G$, the library    files do not  only provide  a
presentation, but, in  addition, a list  of  one or more subgroups  $S_1,
\ldots,  S_r$  of   $G$ such   that   there is  a   faithful  permutation
representation  of $G$ of degree  $\sum_{i=1}^{r} [G:S_i]$ on the set $\{
S_i g \mid 1 \leq i \leq  r, \, g \in  G \}$ of  the cosets of the $S_i$.
This allows to construct the groups as permutation groups, see below. The
`DisplayInformationPerfectGroups'  function  displays  only the available
degree. The message
\begintt
    orbit size = 8
\endtt

in the above example means  that the available permutation representation
is transitive and of degree 8, whereas the message
\begintt
    orbit sizes = 5 + 16
\endtt
means that a nontransitive  permutation representation is available which
acts on two orbits of size 5 and 16 respectively.

The notation   used in the  ``description'' of  a  group  is explained in
section 5.1.2 of \cite{HoltPlesken89}. We  quote the respective page from
there:

{\advance\leftskip by\manindent \advance\rightskip by\manindent
Within a class $Q\,\#\,p$, an isomorphism type  of groups will be denoted
by an  ordered pair of  integers $(r,n)$, where $r \geq  0$ and $n  > 0$.
More precisely, the isomorphism  types in $Q$ \# $p$  of order $p^r  \!\!
\mid  \!\! Q  \!\!  \mid$ will be  denoted by  $(r,1)$, $(r,2)$, $(r,3)$,
$\ldots\,$. Thus $Q$ will always get the size number $(0,1)$.

In  addition to the symbol $(r,n)$,  the groups in $Q  \,  {\sl \#} \, p$
will  also be given a  more descriptive name.  The purpose of  this is to
provide a very rough idea of  the structure of  the group.  The names are
derived in the following manner. First of all, the isomorphism classes of
irreducible  $F_pQ$-modules $M$ with $\mid \!\!  Q  \!\! \mid \mid \!\! M
\!\! \mid \,  \leq 10^6$,  where $F_p$ is   the field of  order $p$,  are
assigned symbols. These  will either  be simply $p^x$,  where $x$  is the
dimension of the module, or, if there is more  than one isomorphism class
of irreducible modules having the same dimension, they will be denoted by
$p^x$, $p^{x^\prime}$,  etc.   The one-dimensional  module   with trivial
$Q$-action  will therefore be denoted  by  $p^1$. These  symbols will  be
listed under the description of $Q$.  The group name consists essentially
of a list  of the composition factors  working from the  top of the group
downwards; hence  it always  starts with the  name  of $Q$ itself.  (This
convention is the most  convenient in our  context,  but it is  different
from that adopted  in the ATLAS (Conway  {\it et al}.~1985), for example,
where composition factors  are listed in the  reverse order. For example,
we denote a  group isomorphic to $SL(2,5)$  by  $A_5 2^1$ rather  than $2
\cdot A_5$.)

Some  other symbols are used  in the name, in order  to give some idea of
the   relationship  between these    composition  factors, and  splitting
properties. We shall now list these additional symbols.
\item{$\times$}  between  two  factors   denotes   a  direct  product  of
      $F_pQ$-modules or groups.
\item{C} (for ``commutator''  between two factors  means  that the second
      lies in the commutator subgroup of the first.  Similarly, a segment
      of the form  $(f_1 \! \times \! f_2) {\sl C} f_3$  would mean  that
      the factors $f_1$ and $f_2$  commute modulo $f_3$ and $f_3$ lies in
      $[f_1,f_2]$.
\item{A} (for ``abelian''  between two factors  indicates that the second
      is  in the  $p$th  power  (but not the commutator subgroup)  of the
      first.   ``A''   may  also   follow  the  factors,   if  bracketed.
\item{E} (for ``elementary abelian''  between two factors  indicates that
      together   they  generate  an  elementary  abelian  group   (modulo
      subsequent factors), but that the resulting $F_pQ$-module extension
      does not split.
\item{N}  (for ``nonsplit''  before  a  factor  indicates  that  $Q$  (or
      possibly its covering group)  splits down as far at this factor but
      not over the factor itself.  So  ``$Q f_1 {\sl N} f_2$'' means that
      the normal subgroup  $f_1f_2$  of the group  has no complement but,
      modulo $f_2$, $f_1$, does have a complement.

Brackets have their obvious meaning. Summarizing, we have:
\item{$\times$} = dirext product;
\item{C} = commutator subgroup;
\item{A} = abelian;
\item{E} = elementary abelian; and
\item{N} = nonsplit.

Here are some examples.
\item{(i)}  $A_5 (2^4 {\sl E} 2^1 {\sl E} 2^4) {\sl A}$  means  that  the
      pairs  $2^4 {\sl E} 2^1$  and $2^1 {\sl E} 2^4$ are both elementary
      abelian of exponent 4.
\item{(ii)}   $A_5 (2^4 {\sl E} 2^1 {\sl A}) {\sl C} 2^1$    means   that
      $O_2(G)$  is of  symplectic type  $2^{1+5}$,  with  Frattini factor
      group  of type   $2^4 {\sl E} 2^1$.   The   ``A''  after the  $2^1$
      indicates that $G$ has a  central cyclic subgroup $2^1 {\sl A} 2^1$
      of order 4.
\item{(iii)}     $L_3(2) ((2^1 {\sl E}) \! \times \! ({\sl N} 2^3 {\sl E}
      2^{3^\prime} {\sl A}) {\sl C}) 2^{3^\prime}$    means   that    the
      $2^{3^\prime}$ factor at the bottom lies in the commutator subgroup
      of the pair $2^3 {\sl E} 2^{3^\prime}$ in the middle, but the lower
      pair $2^{3^\prime} {\sl A} 2^{3^\prime}$  is abelian of exponent 4.
      There  is  also  a submodule  $2^1 {\sl E} 2^{3^\prime}$,  and  the
      covering group  $L_3(2) 2^1$  of  $L_3(2)$  does not split over the
      $2^3$  factor.  (Since $G$ is perfect,  it goes without saying that
      the extension $L_3(2) 2^1$ cannot split itself.)

We  must  stress  that this  notation does   not  always succeed in being
precise  or even unambiguous, and the  reader is free to  ignore it if it
does not seem helpful.\par}

If such a group description has  been given in  the book for $G$ (and, in
fact, this is the case for  most of the library  groups), it is displayed
by the `DisplayInformationPerfectGroups' function. Otherwise the function
provides a less explicit  description  of  the  (in these cases   unique)
Holt-Plesken class to which $G$ belongs, together with a serial number if
this is necessary to make it unique.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Local Variables:
% mode:               text
% mode:               outline-minor
% outline-regexp:     "\\\\Chapter\\|\\\\Section"
% paragraph-start:    "\\\\begin\\|\\\\end\\|\\$\\$\\|.*%\\|^$"
% paragraph-separate: "\\\\begin\\|\\\\end\\|\\$\\$\\|.*%\\|^$"
% fill-column:        73
% End:

