%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%
%W  obsolete.tex             GAP documentation               Frank Lübeck
%%  
%H  @(#)$Id$
%%  

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Chapter{Replaced and Removed Command Names}

\index{obsolete}
\index{deprecated}
\index{legacy}

In general we  try to keep {\GAP}~4 compatible with  former releases as much
as possible. Nevertheless, from time to  time it seems appropriate to remove
some commands  or to change the  names of some commands  or variables. There
are  various reasons  for  that:  Some functionality  was  improved and  got
another (hopefully better) interface, names turned  out to be too special or
too  general for  the underlying  functionality, or  names are  found to  be
unintuitive or inconsistent with other names.


In this  chapter we collect  such old names  while pointing to  the sections
which explain how  to substitute them. Usually, old names  will be available
for several releases;  they may be removed  when they don't seem  to be used
any more.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Section{Group Actions - Name Changes}

\index{group operations}

The  concept  of a  group  action  is sometimes  referred  to  as a  ``group
operation''. In {\GAP}~3  as well as in older versions  of {\GAP}~4 the term
`Operation' was used instead of `Action'.  We decided to change the names to
avoid confusion  with the  term ``operation''  as in  `DeclareOperation' and
``Operations for Xyz''.

Here are some examples of such name changes.

\indextt{Operation}
\indextt{RepresentativeOperation}
\indextt{OperationHomomorphism}
\indextt{FunctionOperation}

\begintt
     OLD                                 NOW USE

     Operation                           Action
     RepresentativeOperation             RepresentativeAction
     OperationHomomorphism               ActionHomomorphism
     FunctionOperation                   FunctionAction
\endtt

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Section{Package Interface - Obsolete Functions and Name Changes}

With {\GAP}~4.4 the package interface was changed. Thereby some functions
became obsolete and the names of some others were made more consistent.

\indextt{DeclarePackage}
\indextt{DeclareAutoPackage}
\indextt{DeclarePackageDocumentation}
\indextt{DeclarePackageAutoDocumentation}
The following functions are no longer needed:
`DeclarePackage', `DeclareAutoPackage', `DeclarePackageDocumentation' and 
`DeclarePackageAutoDocumentation'. They are substituted by entries in the 
`PackageInfo.g' files, see "Ext:The PackageInfo.g File". 

The following function names were changed.

\indextt{RequirePackage}
\indextt{ReadPkg}
\indextt{RereadPkg}
\indextt{CreateCompletionFilesPkg}
\begintt
     OLD                                 NOW USE

     RequirePackage                      LoadPackage
     ReadPkg                             ReadPackage
     RereadPkg                           RereadPackage
     CreateCompletionFilesPkg            CreateCompletionFilesPackage
\endtt

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Section{Normal Forms of Integer Matrices - Name Changes}

\index{Smith normal form}
\index{Hermite normal form}

Former versions of  {\GAP}~4 documented several functions  for computing the
Smith or  Hermite normal form of  integer matrices. Some of  them were never
implemented  and it  was unclear  which commands  to use.  The functionality
of  all  of   these  commands  is  now   available  with  `NormalFormIntMat'
(see~"NormalFormIntMat") and a few interface functions.


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Section{Miscellaneous Name Changes or Removed Names}

\indextt{QUIET}
\indextt{BANNER}
\indextt{GAPInfo}

In former releases  of {\GAP}~4 there were some global  variable names bound
to general  information about the  running {\GAP} (path names,  command line
options, ...). Although  they were not officially documented  they were used
by several users and in some packages. We mention here `BANNER' and `QUIET'.
This type  of information is now  collected in a record  with name `GAPInfo'
and will become documented after a test phase.


Here are some further name changes.

\indextt{MonomialTotalDegreeLess}
\indextt{NormedVectors}
\begintt
     OLD                                 NOW USE

     MonomialTotalDegreeLess             MonomialExtGrlexLess
     NormedVectors                       NormedRowVectors
\endtt

