\documentclass{article}
\usepackage{cd-cover}
%\usepackage{syntax}\renewcommand{\rmdefault}{esx}
%\def\bs{$\backslash$}
\begin{document}
\bookletsheet{
{\Large Installation}
\medskip

The installation of \textsf{GAP} 4.3 is explained in the files 
\texttt{ftp/gap4/README.first}
and, depending on your operating system, \texttt{ftp/gap4/INSTALL...}
\medskip

\textsf{GAP} 4.3 can be installed on UNIX/Linux, Windows and 
Macintosh computers. Particular hints and an installation script for 
UNIX can be found in \texttt{ftp/gap4/README.UNIX} and 
\texttt{ftp/gap4/inst\_all.unix}.
\medskip

All archives are available in \texttt{.zoo}-, \texttt{.zip}- and
\texttt{.tar.gz}-format.
\medskip

In case you also want to install {\sf GAP} 3, you can use the archives in the
{\tt ftp/gap-3.4.4} subdirectory.
\medskip

After you have installed {\sf GAP}, please send us an email at
\hbox{\tt gap@dcs.st-and.ac.uk}, telling us that you installed {\sf GAP}.
\medskip

Please lend this CD to friends and colleagues who also want to install
{\sf GAP}. This package is distributed under the terms of the GNU Public
License (GPL). Details of the license and copyright are contained in the
distribution.
\medskip

The CD also contains a copy of the \textsf{GAP} Website, the start page is
in \texttt{WWW/index.html}.
\bigskip

The GAP Group
}{
\begin{center}
\textsf{\Huge\bf GAP 4.3}
\bigskip

\mbox{}
\bigskip

{\bf\Large Groups, Algorithms and Programming}
\bigskip

{\large A System for Computational Algebra}
\bigskip

\mbox{}
\bigskip

May 10, 2002
\end{center}
}

\backsheet{\textsf{\bf GAP 4.3}, May 10, 2002}{
\begin{center}
\textsf{\LARGE\bf GAP 4.3}
\medskip

Groups, Algorithms and Programming
\medskip

May 10, 2002
\end{center}

{\Large Contains:}
\begin{description}
\item GAP 4.3: Latest distribution version. No bugfixes at time of printing.
\item \texttt{ftp.gap-system.org}: part of the Ftp server image.
\item \texttt{www.gap-system.org}: Web server image.
\end{description}
}
\end{document}
