\documentclass{article}
\usepackage{cd-cover}
\usepackage{syntax}\renewcommand{\rmdefault}{esx}
\def\bs{$\backslash$}
\begin{document}
\bookletsheet{
\begin{center}
{\Large Installation}
\medskip\\
\end{center}
\begin{raggedright}
Complete installation instructions can be found in the
{\tt INSTALL} files in the {\tt ftp/gap4} directory. Get the appropriate
files from this directory and install from the archives. Make sure you get
all bugfixes (and check at
\hbox{\tt http:$/\!/$www-gap.dcs.st-and.ac.uk/gap/Info4/bugfixes.html}
whether there are any new bugfixes).
\\
Macintosh users please see the READMAC.TXT file in the root directory!
\smallskip\\
{\small
(It is possible, but a bit more complicated to run {\sf GAP} off the CD:
Create a {\sf GAP} directory on your harddisk (this is the directory to
which you would install all further bugfixes). Copy the {\tt src}, {\tt cnf}
directories and the configuration files there and compile {\sf GAP} as
usual. To start {\sf GAP}, give the CD as a second {\sf GAP} root directory via
the {\tt -l} command line option. See section~``GAP Root Directory'' in
chapter~``Files and Filenames'' of the reference manual.\\ Windows 9x users
can copy and edit the \hbox{\tt gap4r1{\bs}bin{\bs}gap.bat} file.)
}
\bigskip\\
In case you also want to install {\sf GAP} 3, you can use the archives in the
{\tt ftp/gap-3.4.4} subdirectory.
\bigskip\\
After you have installed {\sf GAP}, please send us an email at
\hbox{\tt gap@dcs.st-and.ac.uk}, telling us that you installed {\sf GAP}.
\bigskip\\
Please lend this CD to friends and colleagues who also want to install
{\sf GAP}.
\end{raggedright}
}{
\begin{center}
{\Huge\bf GAP}\bigskip\bigskip\\
{\bf\Large Groups, Algorithms and Programming}\bigskip\\
{\large A System for Computational Algebra}\bigskip\bigskip\\
15-December-1999
\end{center}
}
\backsheet{{\bf GAP}, 15-December-1999}{
\begin{center}
{\LARGE\bf GAP}\medskip\\
Groups, Algorithms and Programming\medskip\\
15-December-1999
\end{center}
{\Large Contains:}
\begin{description}
\item GAP 4r1: GAP 4.1 with bugfix 7.
\item ftp-gap.dcs.st-and.ac.uk: Ftp server image.
\item www-gap.dcs.st-and.ac.uk: Web server image.
\end{description}
{\LARGE\bf\hphantom{F}{GAP} group}\medskip\\
%{\huge\bf\hphantom{II}}\begin{minipage}{105mm}
School of Mathematical and Computational Sciences, University of St~Andrews,\\
The North Haugh, St~Andrews, Fife KY16 9SS, United Kingdom\medskip\\
\begin{tabular}{l@{}l}
&{http://www-gap.dcs.st-and.ac.uk/{\tt\char126}gap}\smallskip\\
%{\large\fontfamily{dbs}\selectfont v}
&{gap@dcs.st-and.ac.uk}\\
&{gap-trouble@dcs.st-and.ac.uk}\smallskip\\
&(For any problems loading or running {\sf GAP}, please use the latter
address.)\medskip\\
{\large Tel. }\qquad&
++44(1334)463251\qquad\quad
{\large Fax. }
++44(1334)463278\\
\end{tabular}
%\end{minipage}
}
\end{document}
