%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%
%W  opers.tex                 GAP documentation            Heiko Theissen
%%
%H  @(#)$Id$
%%
%Y  Copyright 1997,  Lehrstuhl D fuer Mathematik,  RWTH Aachen,   Germany
%%
%%  This file contains a tutorial introduction to operations.
%%


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Chapter{Attributes and their Methods}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Section{Attributes}

In the  preceding chapters, we have  seen how to obtain information about
mathematical objects in {\GAP}: We have to pass the object as an argument
to a function, e.g., `Size( <G> )', and the function will return a value,
in our example an integer  which is the  size of the group <G>. Computing
the size  of a  group generally  requires  a substantial amount  of work,
therefore it seems desirable to store the size somewhere once it has been
calculated. You should imagine that {\GAP} stores  the size in some place
associated  with the object  <G> when `Size( <G>  )'  is executed for the
first time, and if this function  call is executed  again later, the size
is simply looked up and returned, without further computation.

\index{getter!of an attribute}\index{setter!of an attribute}%
\index{tester!of an attribute}\index{methods}%
This means that the  behavior  of the  function  `Size' has to  depend on
whether the size for the argument <G> is already known,  and if not, that
the size  must be  stored after it  has been  calculated. These two extra
tasks are  done  by two other   functions that accompany `Size(   <G> )',
namely the *tester*  `Tester( Size )( <G>  )'  and the  *setter* `Setter(
Size )( <G>, <size> )'. The tester returns `true' or `false' according to
whether <G>  already knows its  size, and  the setter  puts <size> into a
place from where <G> can directly look  it up. The function `Size' itself
is called the  *getter*, and from the  preceding  discussion we see  that
there must really be two getters, or rather two *methods* for the getter:
The  first method is   used when the   tester returns `false';  it is the
method which first does the real computation and then executes the setter
with   the computed value.  The second  method is  used   when the tester
returns `true'; it simply returns the stored value. This second method is
also called the *system getter*.
\beginexample
    gap> G := Group( (1,2,3,4,5,6,7,8), (1,2) );; Size( G ); time;
    40320
    3583   # this may of course vary on your machine
    gap> Size( G ); time;
    40320
    0
\endexample
The convenient thing  for the user  is that  {\GAP} automatically chooses
the right method  for the getter, i.e.,  it calls the real-work getter at
most once  and the system getter  in all subsequent occurrences. *At most
once* because the value of a function call like `Size( <G> )' can also be
set for <G>  before the getter  is called at all; to  do this you  simply
call the setter directly. By the way: for  most attributes the tester has
an alias like `HasSize' for `Tester( Size )' and the setter is also known
as `SetSize'.

Every *attribute* in  {\GAP} is represented by  a triple  of a getter,  a
setter   and a  tester.  When a  new  attribute  is  declared, all  three
functions are created together and  the getter contains references to the
other two.  This is necessary because when  the getter is called, it must
first  consult the tester,   and perhaps execute the  setter  in the end.
Therefore the getter could be implemented as follows:
\begintt
    getter := function( obj )
    local   value;
|_
        if tester( obj )  then
            value := system_getter( obj );
|_
        else
            value := real_work_getter( obj );
            setter( obj, value );
        fi;
|_
        return value;
    end;
\endtt
The  only  function which  depends on   the  mathematical  nature  of the
attribute  is  the  real-work getter,  and this   is  of course  what the
programmer of  an  attribute has to  install.  In both cases,  the getter
returns  the same value, which  we also call  the  value of the attribute
(properly: the value of the attribute for the object `obj').

\exercise  What  do you think:  Is  the {\GAP} function  `Random' also an
attribute?

\answer Of course  not, because then the  first random element of a group
would be stored by  the setter and returned over  and  over again by  the
system getter every  time `Random' is  called in  the sequel.  Not a very
random behavior.

There is a general important rule about attributes: *Once the value of an
attribute for an object has been set, it cannot be reset, i.e., it cannot
be changed any more.* This is achieved by having two methods not only for
the getter but also for the setter: If an object already has an attribute
value stored, i.e., if the tester  returns `true', the setter simply does
nothing.
\beginexample
    gap> SetSize( G, 0 ); Size( G );
    40320
\endexample

\exercise Experiment  with {\GAP} to  find out whether immutability of an
object prevents a setter from setting a previously unset value.

\answer It does  not. That an  object like a  group is constant  does not
mean  that   additional information cannot    be entered. Such additional
information does not change the mathematical identity of the object.

{\bf  Summary.} In this section  we have introduced attributes as triples
of getter, setter   and tester and    we have explained how  these  three
functions work together behind  the  scenes to provide automatic  storage
and look-up of  values that have once been  calculated. We have seen that
there  are  several  methods for  the   same function  among which {\GAP}
automatically selects an appropriate one.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Section{Properties and filters}\null

\atindex{properties}{|indexit}\atindex{filters}{|indexit}%
\index{methods!selection}%
Certain attributes, like `IsAbelian', are boolean-valued. Such attributes
are known to {\GAP} as *properties*, because their values are stored in a
slightly different way.   A property also  has a  getter, a  setter and a
tester, but  in this case,  the  getter as well  as the  tester returns a
boolean value. Therefore {\GAP}  stores   both values  in the same   way,
namely as bits  in a boolean  list, thereby treating property getters and
all testers (of attributes or properties) uniformly. These boolean-valued
functions  are called  *filters*.  You can imagine  a filter  as a switch
which is  set either  to `true' or  to `false'.  For every  {\GAP} object
there is a boolean list which has reserved a  bit for every filter {\GAP}
knows  about. Strictly speaking, there   is one bit for every  *simple
filter*, and these  simple filters can be  combined with `and' to form
other filters (which are then `true' if and only if all the corresponding
bits are set to `true').
For example, the filter `IsPermGroup and IsSolvableGroup' is made up from
several simple filters.

Since they allow only two values, the bits which represent filters can be
compared very quickly, and the scheme by which {\GAP} chooses the method,
e.g., for a getter or a setter (as we have seen in the previous section),
is mostly based  on the examiniation of filters,  not on the examiniation
of other  attribute   values. Details  of   this *method selection*   are
described  in chapter~"prg:chapter  Method  Selection" of the programming
manual.

We  only present the following  rule  of the  thumb here: Each  installed
method for an attribute, say `Size(  <obj> )', has a ``required filter'',
which is made  up from certain simple filters  which must yield `true'
for <obj> if this method is to be applicable. To execute a call of `Size(
<obj>   )', {\GAP} selects  the  applicable  method whose required filter
combines the most  simple filters because  it thinks that the more  an
algorithm requires of <obj>, the more efficient  it will be. For example,
if <obj>  is a permutation group  that is not  (known to be)  solvable, a
method  with required filter   `IsPermGroup  and IsSolvableGroup'  is not
applicable,  hence a  method with  required  filter `IsPermGroup' will be
chosen. On  other  hand, if <obj>  was  known to be   solvable, the first
method with required filter  `IsPermGroup  and IsSolvableGroup' would  be
selected. It may happen that a  method fails although the required filter
is  `true' for the  argument. In such  cases, the method will execute the
statement  `TryNextMethod();' then the  next   applicable method will  be
chosen.

\indextt{TryNextMethod}%
\exercise How  can  `TryNextMethod()' be used   to exploit  the potential
solvability of a permutation group?

\answer Suppose  there is a method for  solvable permutation groups which
is far superior to the general method and which  is able to return `fail'
if the  group is not solvable  (rather than end up  in an infinite loop).
Then one can install a  method with  required filter `IsPermGroup'  which
looks like
\begintt
    function( G )
    local  value;
        value := solvable_method( G );
        if value <> fail  then  return value;
                          else  TryNextMethod();  fi;
    end;
\endtt
This method would then be tried on every permutation group (whether known
to be  solvable or  not),  and it would  include a  mandatory solvability
test.  {\GAP} pursues  such  a strategy for   `Size',  using the solvable
method from \cite{Sims90b}.

\edef\solsimsex{\the\chapno.\the\exno}%
If no applicable method  (or no next applicable  method) is found, {\GAP}
stops with an error message either like
\begintt
    Error argument for 'Size' must have categories '
    [ "IsListOrCollection" ]' at ...
\endtt
or like
\begintt
    Error no method found for operation 'Size' with 1 argument at ...
\endtt
The  first variant occurs  if the  argument to   `Size' was of  the wrong
category, e.g., if you asked for `Size( 1 )'. Every {\GAP} object belongs
to  a  number of     categories,   e.g., `IsInt(   1    ) =  true',   but
`IsListOrCollection( 1 ) = false', and  the latter is required by `Size'.
The  second variant   occurs if the   category  was right, e.g.,   if the
argument  was a list  or collection,  but there  was no applicable method
installed. If these remarks sound rather mysterious  to you, you can find
the details in  sections~"prg:Categories" and~"prg:Applicable Methods and
Method Selection" in the programming manual.

{\bf  Summary.}  In   this section  we  have   specialized  attributes to
properties and then generalized them again to  filters. The values of the
filters for an object has govern how the  object is treated in the method
selection.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Section{Immediate and true methods}\null

\index{methods!immediate}\index{methods!true}%
In exercise~\solsimsex\ we have mentioned that the operation `Size' has a
method  for solvable permutation  groups that is  so  far superior to the
method for  general permutation groups that  it seems worthwile to try it
even if nothing  is  known about solvability   of the group of which  the
`Size' is to   be  calculated. There are   other  examples where  certain
methods  are even ``cheaper'' to  execute. For example,  if the size of a
group is known  it is easy to check  whether  it is  odd, and  if so, the
Feit-Thomposon  theorem allows us to set  `IsSolvableGroup' to `true' for
this group.   {\GAP} utilizes   this  celebrated  theorem  by  having  an
*immediate  method* for `IsSolvableGroup'  with required filter `HasSize'
which checks parity of the size and either sets `IsSolvableGroup' or does
nothing, i.e.,   calls `TryNextMethod()'.   These immediate  methods  are
executed  automatically for an  object  whenever the   value of a  filter
changes, so solvability of a group will automatically be detected when an
odd size has been calculated for it (and therefore the value of `HasSize'
for that group has changed to `true').

How immediate methods  are  installed   is described in    "prg:Immediate
Methods" in the programming manual.

Some methods are  even more immediate,   because they do not  require any
calculation  at all: They  allow a filter to  be set if another filter is
also set. In other words,   they  model a mathematical implication   like
`IsGroup   and   IsCyclic    $\Rightarrow$   IsSolvableGroup'  and   such
implications  can be installed  in {\GAP}  as  *true methods*. To execute
true methods, {\GAP} only needs to do  some bookkeeping with its filters,
therefore true methods are much faster than immediate methods.

How true methods are installed is described in "prg:Logical Implications"
in the programming manual.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Section{Operations and method selection}\null

\index{operations}%
The method selection  is not only  used to  select  methods for attribute
getters but also for arbitrary *operations*, which can have more than one
argument.  In this case,   there is a   required filter for each argument
(which must yield `true'  for the corresponding  arguments), and a method
may  also  require  a certain   relation   between the families  of   the
arguments. For example, the methods for  `ConjugateGroup( <grp>, <elm> )'
require that <elm>  lies in the family   of elements from  which <grp> is
made, i.e., that  the family of  <elm> equals the ``elements family''  of
<grp>.   For  permutation groups,  the  situation   is   quite easy:  all
permutations form  one family, `PermutationsFamily', and  all permutation
groups belong to the `CollectionsFamily( PermutationsFamily )' (and so do
all    sets of permutations, e.g.,     cosets of permutation groups). For
further information, see "prg:Families" in the programming manual.
\beginexample
    gap> ElementsFamily( CollectionsFamily( PermutationsFamily ) );
    NewFamily( "PermutationsFamily", [ 27 ],
               [ 27, 86, 89, 92, 96, 100, 106, 111 ] )
    gap> last = PermutationsFamily;
    true
\endexample

\exercise Why does {\GAP} refuse to compute `One( FreeGroup( 1 ) ) * One(
FreeGroup(  1   )   )'? (Look   at  the   family   relation  required for
multiplication.)

\answer Every call of  `FreeGroup' constructs a new  family of free group
elements, so both `One's belong   to different families. The methods  for
multiplication, however, require that the families of their arguments are
the same.

\indextt{KnownPropertiesOfObject}\indextt{KnownTruePropertiesOfObject}%
\indextt{KnownAttributesOfObject}%
If  you want to know   which properties are   already known for an object
<obj>,  or  which properties  are  known  to  be true,   you can use  the
functions        `KnownPropertiesOfObject(     <obj>    )'         resp.\
`KnownTruePropertiesOfObject( <obj> )'. This will  print a list of  names
of  properties.  These names are  also   the identifiers  of the property
getters, by which  you can  retrieve  the value  of  the properties  (and
confirm that they are really `true').  Analogously, there is the function
`KnownAttributesOfObject' which lists the names  of the known attributes,
leaving out the properties.

\indextt{ApplicableMethod}%
Since {\GAP} lets you  know what it already  knows about an object, it is
only natural  that   it also  lets   you know what  methods  it considers
applicable for a certain method, and in  what order it  will try them (in
case `TryNextMethod()'  occurs).   `ApplicableMethod( <opr>, [   $arg_1$,
$arg_2$, \dots\ ]  )' returns the  first  applicable method for the  call
`<opr>( $arg_1$, $arg_2$,  \dots\ )'. More  generally, `ApplicableMethod(
<opr>, [ \dots\ ], 0, <nr> )' returns the <nr>th applicable method (i.e.,
the one  that would be  chosen  after  $<nr>-1$ `TryNextMethod's) and  if
`<nr>  = "all"', the sorted list  of  all applicable methods is returned.
For  details,  see "prg:Applicable Methods  and  Method Selection" in the
programming manual.

\indextt{TraceMethods}%
If you want to see which methods  are chosen for certain operations while
{\GAP}  code is being executed,  you can call the function `TraceMethods'
with a list of these operations as arguments.
\beginexample
    gap> TraceMethods( [ Size ] );
    gap> g:= Group( (1,2,3), (1,2) );;  Size( g );
    |#I  Size: method for a permutation group
    |#I  Setter(Size): system setter
    |#I  Size: system getter
    |#I  Size: system getter
    6
\endexample
The system getter is called once to fetch  the freshly computed value for
returning  to the user.  The  second  call is  triggered by  an immediate
method. To  find out  by which,  we can trace   the immediate  methods by
saying `TraceImmediateMethods( true )'.
\beginexample
    gap> TraceImmediateMethods( true );
    gap> g:= Group( (1,2,3), (1,2) );;
    |#I  immediate: IsFinitelyGeneratedGroup
|_
    gap> Size( g );
    |#I  Size: method for a permutation group
    |#I  immediate: IsNonTrivial
    |#I  immediate: IsNonTrivial
    |#I  immediate: IsFinitelyGeneratedGroup
    |#I  immediate: IsCyclic
    |#I  immediate: IsFinitelyGeneratedGroup
    |#I  immediate: IsNonTrivial
|_
    |#I  Setter(Size): system setter
    |#I  Size: system getter
    |#I  immediate: IsPerfectGroup
    |#I  Size: system getter
    |#I  immediate: IsEmpty
    6
|_
    gap> TraceImmediateMethods( false );
    gap> UntraceMethods( [ Size ] );
\endexample
The last two lines switch off tracing  again. We now  see that the system
getter was called by the immediate method for `IsPerfectGroup'. The group
`g' was   known to be finitely   generated  already before the   size was
computed,   therefore the   calls     of   the  immediate    method   for
`IsFinitelyGeneratedGroup' after the  call of `Size' have other arguments
than~`g'. Also the above-mentioned immediate method for `IsSolvableGroup'
was not used because the solvability of `g'  was already found out during
the size calculation (cf. exercise~\solsimsex).

{\bf Summary.}  In this section  and the  last we have  looked some  more
behind the  scenes and seen  that {\GAP} automatically executes immediate
and true  methods  to deduce  information about  objects  that is cheaply
available.  We  have seen how   this  can be  supervised  by tracing  the
methods.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Local Variables:
% mode:               text
% mode:               outline-minor
% outline-regexp:     "\\\\Chapter\\|\\\\Section\\|\\\\stars"
% paragraph-start:    "\\\\begin\\|\\\\end\\|\\$\\$\\|.*%\\|^$"
% paragraph-separate: "\\\\begin\\|\\\\end\\|\\$\\$\\|.*%\\|^$"
% fill-column:        73
% End:

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%
%E  opers.tex . . . . . . . . . . . . . . . . . . . . . . . . . ends here
