
\Chapter{Introduction}

This package contains methods to compute with finite Lie algebras; that
is, with Lie algebras over finite fields. The functions in this package 
are often particularly effective if the considered Lie algebras are 
solvable, since there are various special methods for finite solvable 
Lie algebras implemented.

The GAP system contains various methods to compute with Lie algebras,
see \cite{DGr00} for background. This package is based on these methods 
and extends them in the case of a finite Lie algebra, see \cite{Eic04a}
for a description of some of the methods.

\Chapter{Solvable Lie algebras}

First, we describe some of the basic functions for finite solvable Lie 
algebras which are used throughout the package.

\> LowerNilpotentBasis(<L>)

Takes as input a solvable Lie algebra <L> and returns a record with
entries <basis> and <weights>. The entry <basis> is a list of elements
in <L> which forms a basis. The basis exhibits the lower nilpotent series 
of <L> which is a automorphism-invariant series with abelian factors. The 
entry <weights> is a list which describes the steps of the series.

\> UpperNilpotentBasis(<L>)

Takes as input a solvable Lie algebra <L> and returns a record with
entries <basis> and <weights>. The entry <basis> is a list of elements
in <L> which forms a basis. The basis exhibits the upper nilpotent series 
of <L> which is a automorphism-invariant series with abelian factors. The 
entry <weights> is a list which describes the steps of the series.

\Chapter{Cohomology and extensions for finite Lie algebras}

\> LieOneCobounds( <L>, <B>, <M> )
\> LieOneCocycles( <L>, <B>, <M> )
\> LieOneCohomology( <L>, <B>, <M> )

These functions take as input a Lie algebra <L>, a basis <B> of <L> 
and a list of matrices <M> such that the map mapping the elements of
<B> to the elements of <M> induces a representation of <L>.

Let $B = \{b_1, \ldots, b_n\}$ and let $V$ be the natural $L$-module
defined by the given representation. Then the map $Z^1(L, V) \rightarrow 
V^n : \varphi \mapsto (b_1^\varphi, \ldots, b_n^\varphi)$ is a faithful
representation of $Z^1(L, V)$.

The functions LieOneCobounds and LieOneCocyles return lists of vectors
which form a basis for the images of the representations of $Z^1(L, V)$ 
and $B^1(L, V)$, respectively.

The function LieOneCohomology returns a linear map onto the factor of
the results of LieOneCocycles modulo LieOneCobounds. Thus the image
of this map is a faithful representation of $H^1(L, V)$.

\> LieTwoCobounds( <L>, <B>, <M> )
\> LieTwoCocycles( <L>, <B>, <M> )
\> LieTwoCohomology( <L>, <B>, <M> )

These functions take as input a Lie algebra <L>, a basis <B> of <L> 
and a list of matrices <M> such that the map mapping the elements of
<B> to the elements of <M> induces a representation of <L>.

The functions LieTwoCobounds and LieTwoCocyles return lists of vectors
which form a basis for the images of representations of $Z^2(L, V)$ 
and $B^2(L, V)$, respectively. The chosen representations may not be
faithful!

The function LieTwoCohomology returns a linear map onto the factor of
the results of LieTwoCocycles modulo LieTwoCobounds. The image of this
map is a faithful representation of $H^2(L, V)$.

\> LieExtensionByCocycle( <L>, <B>, <M>, <c> )

This function takes as input a Lie algebra <L>, a basis <B> of <L>,
a list of matrices <M> such that the map mapping the elements of <B> 
to the elements of <M> induces a representation of <L> and a 2-cocyle
<c> in the representation $\psi$ of $Z^2(L, V)$.

The function returns a Lie algebra which is an extension of <L> by
the natural <L>-module $V$ defined by the representation via <c>.

\Chapter{Automorphism groups and isomorphism testing}

\> AutomorphismGroupOfSolvableLieAlgebraBySpecial(<L>)
\> AutomorphismGroupOfSolvableLieAlgebraByFitting(<L>)
\> AutomorphismGroupOfSolvableLieAlgebra(<L>)
\> AutomorphismGroupOfLieAlgebra(<L>)

All of these functions take as input a finite Lie algebra <L>. In the first
three cases <L> must be solvable. All functions return the automorphism
group of <L>. The last function can be really slow if <L> is not solvable.

\> AutomorphismGroup(<L>) A

Computes the automorphism group of the finite Lie algebra <L> and stores 
it as an attribute.

\> CanonicalFormOfSolvableLieAlgebra(<L>) A

Takes as input a finite solvable Lie algebra <L> and it returns another 
finite solvable Lie algebra $K$ isomorphism to <L> whose structure 
constants are in a canonical form. Two Lie algebras <L> and <H> are 
isomorphic if and only if the structure constants of their canonical 
forms are equal. The computed canonical form is stored as an attribute.

\> AreIsomorphicLieAlgebras(<L>, <H>) M

This function checks whether the finite Lie algebras <L> and <H> are
isomorphic. The function can be really slow if <L> or <H> are not
solvable.


