%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%
%W  ctbldiff.tex                                                Thomas Breuer
%%
%H  @(#)$Id$
%%
%Y  Copyright (C)  1999,  Lehrstuhl D fuer Mathematik,  RWTH Aachen,  Germany
%Y  (C) 1999 School Math and Comp. Sci., University of St.  Andrews, Scotland
%%
%%  Plain TeX file, format copied from Simon Norton's famous lists of
%%  ATLAS Improvements
%%

\parskip 1ex plus 0.5ex minus 0.5ex
\parindent0pt

\font\sf=cmss10
\def\GAP{{\sf GAP}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\centerline {\bf Changes in the {\GAP} Character Table Library}

\bigbreak

This list contains the changes in the {\GAP} character table library
since the official upgrade for {\GAP}~3.4 in October 1996.
We denote mathematical errors by {\bf ***} and new information
by {\bf NEW}\null.
We use {\bf C} to denote changes that are not obviously corrections;
the number of these changes is kept small.

\bigbreak

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
{\bf Release of {\GAP}~4.1 in July 1999}

\bigbreak

{\bf Brauer Tables}

Changes are assigned to the simple group involved,
and shown in alphabetical order.

\medbreak

\halign{\hfil{\bf #}&\quad$#$\hfil&\quad#\hfil$\;\>$&
 \vtop{\parindent=0pt\hsize=38em\strut#\strut}\cr
***&^2E_6(2)&:&The faithful characters of $2.{}^2E_6(2)$ and
$2.{}^2E_6(2).2$ mod $19$ were corrected.\cr
% (by J"urgen M"uller)
NEW&A_{13}&:&Indicators of $A_{13}$ and $S_{13}$ mod $2$ are now known.
\cr
NEW&A_{14}&:&The tables of $A_{14}$ mod $2$, $11$, $13$ and
tables of $S_{14}$ mod $3$, $5$, $7$ are now known.\cr
NEW&A_{15}&:&All Brauer tables of $S_{15}$ are now known.\cr
% (by J"urgen M"uller)
NEW&A_{16}&:&All Brauer tables of $S_{16}$ are now known.\cr
% (by J"urgen M"uller)
NEW&A_{17}&:&All Brauer tables except the $3$-modular one of $S_{17}$
are now known.\cr
% (by J"urgen M"uller)
NEW&Co_2&:&The degree $156\,538$ character of $Co_2$ mod $2$ is now
proved.\cr
NEW&Co_3&:&One more indicator of $Co_3$ mod $2$ is now known.\cr
***&Fi_{22}&:&The faithful characters of $6.Fi_{22}$ and $6.Fi_{22}.2$
mod $5$
were corrected.\cr
***&L_3(4)&:&The faithful characters of $12_2.L_3(4)$ mod $7$ were
corrected.\cr
NEW&O_8^+(3)&:&The degree $50\,596$ characters of $O_8^+(3)$ mod $2$
are now proved.
Consequently, also the degree $101\,192$ of $O_8^+(3).2_1$ mod $2$,
the degrees $50\,596$ and $101\,192$ of $O_8^+(3).2_2$ mod $2$,
the degrees $50\,596$ and $151\,288$ of $O_8^+(3).3$ mod $2$,
and the degree $202\,384$ of $O_8^+(3).4$ mod $2$ are now proved.\cr
C&ON&:&The tables of $ON$ and $ON.2$ mod $19$ were changed
in order to respect the choice of classes in Rob Wilson's
``Atlas of Representations''.\cr
NEW&ON&:&The tables of $ON$ mod $11$ and mod $31$ are now known,
% (computed by Markus Ottensmann)
as well as two new indicator values for $ON$ mod $2$.\cr
C&Ru&:&The tables of $Ru$ and $2.Ru$ mod $5$ and mod $7$ were changed
in order to respect the choice of classes in Rob Wilson's
``Atlas of Representations''.\cr
NEW&Ru&:&The tables of $Ru$ and $2.Ru$ mod $13$ and $29$ are now known,
% (computed by Frank Roehr)
as well as all indicator values of $Ru$ mod $2$.\cr
***&S_6(3)&:&The characters of $S_6(3)$ and $2.S_6(3)$ were corrected.
(The changes do not affect the tables of $S_6(3).2$ and $2.S_6(3).2$
mod $7$.)\cr
% (by J"urgen M"uller)
***&Suz&:&The faithful characters of $6.Suz$ and $6.Suz.2$ mod $7$ were
corrected.\cr
% (bug reported by William Husen, correction by J"urgen M"uller)
NEW&Th&:&The table of $Th$ mod $19$ is now known.\cr}

\bigbreak

{\bf Ordinary Tables}

The following changes affect several ordinary tables.

\medbreak

\halign{\hfil{\bf #}&\quad
 \vtop{\parindent=0pt\hsize=43em\strut#\strut}\cr
C&Whitespace at the end of {\tt InfoText} strings was removed.\cr
NEW&Various class fusions were added.\cr
NEW&Components {\tt tomidentifier} and {\tt tomfusion} were added
in order to provide a (preliminary) interface to
the library of tables of marks.\cr
C&In the library tables of alternating and symmetric groups,
the {\tt classtext} components
(partitions parametrizing the conjugacy classes;
in some cases, this had been hidden inside the {\tt CAS}
component of the table)
were replaced by values of the attribute {\tt ClassParameters}.\cr
NEW&The tables of $L_2(q)$ were added for those values of $q$ for which
the table of marks of $L_2(q)$ is now contained in the {\GAP} library.\cr
NEW&In the library tables of symmetric groups,
the partitions parametrizing the irreducible characters
are stored on the tables,
as value of the attribute {\tt CharacterParameters}.\cr
C&The {\tt Identifier} values of a few tables have been changed.
For example, the table of {\tt L4(3).2\^{}2} was previously known only as
{\tt psl(4,3).v4}.
The old names are still valid.\cr
***&The character tables with identifiers {\tt iu332}, {\tt D2MJ4},
and {\tt P4L82} were removed.
The former two tables were incomplete, the latter one was wrong.\cr
NEW&The ordinary tables of all maximal subgroups
(and their class fusions)
are now available for the groups $G_2(3)$, $J_3.2$, $2.M_{12}$,
$M_{12}.2$, $M_{22}.2$, and $O_8^+(3)$.\cr}

\medbreak

The following changes are assigned to specific simple groups,
and shown in alphabetical order.

\medbreak

\halign{\hfil{\bf #}&\quad$#$\hfil&\quad#\hfil$\;\>$&
 \vtop{\parindent=0pt\hsize=38em\strut#\strut}\cr
***&A_6&:&The table automorphisms of $4.A_6.2_3$ were corrected.\cr
NEW&Fi_{22}&:&The table of the maximal subgroup $2^7\!:\!S_6(2)$
of $Fi_{22}.2$ was added.\cr
% (computed by E. Mpono)
NEW&Fi_{22}&:&The table of the maximal subgroup $2^6\!:\!U_4(2).2$
of the maximal subgroup $2^6\!:\!S_6(2)$ of $Fi_{22}$ was added.\cr
% (computed by E. Mpono)
NEW&HS&:&The tables (and fusions) of several normalizers of chains of
$p$-subgroups were added.\cr
C&J_4&:&The classes and the characters of the maximal subgroup of type
$2^{10}\!:\!L_5(2)$ were reordered,
and the identifier was changed from {\tt l52m10} (from the CAS library)
to {\tt 2\^{}10:L5(2)}.\cr
% Make the old table available again, with the old name! (ctomaxi6.tbl)
NEW&McL&:&The table of the seventh maximal subgroup of $McL.2$ was added.
\cr
C&O_8^+(2)&:&The classes and the characters of the maximal subgroup of
type $2^6\!:\!A_8$ were reordered,
and the identifier was changed from {\tt mo81p} (from the CAS library) to
{\tt 2\^{}6:A8}.\cr
% Make the old table available again, with the old name! (ctomisc2.tbl)
% The same holds for the table with name `y' ! (ctomisc6.tbl)
C&O_8^+(3)&:&The fusions from $O_7(3)$ and $3^6\!:\!L_4(3)$ were changed
to the ones listed in the Atlas of Finite Groups.\cr
NEW&O_{10}^+(2)&:&The table of the maximal subgroup $2^8\!:\!O_8^+(2)$
was added.\cr
NEW&S_{10}(2)&:&The table of the subgroup $2^8\!:\!S_8(2)$ was added.\cr
NEW&U_4(3)&:&The tables of $2.U_4(3).(2^2)_{122}$ and $6_2.U_4(3).2_3'$
were added.\cr}

\bigbreak

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
{\bf Release of {\GAP}~4.2 in February 2000}

\bigbreak

{\bf Brauer Tables}

Changes are assigned to the simple group involved,
and shown in alphabetical order.

\medbreak

\halign{\hfil{\bf #}&\quad$#$\hfil&\quad#\hfil$\;\>$&
 \vtop{\parindent=0pt\hsize=38em\strut#\strut}\cr
NEW&A_{14}&:&Table of $S_{14}$ mod $2$ is now known.\cr
% (first computed by Dave Benson, added by J"urgen M"uller)
***&A_{16}&:&Corrected principal block of the table of $S_{16}$ mod $2$.\cr
NEW&ON&:&The tables of $3.ON$ mod $11$ and $31$ are now known.\cr
C&ON&:&The tables of $3.ON$ and $3.ON.2$ mod $19$ were changed
in order to respect the choice of classes in Rob Wilson's
``Atlas of Representations''.\cr}
% (concerning only irreducibles of $3.ON$ of degrees $45090$ and $77670$)

\bigbreak

{\bf Ordinary Tables}

The following changes affect several ordinary tables.

\medbreak

\halign{\hfil{\bf #}&\quad
 \vtop{\parindent=0pt\hsize=43em\strut#\strut}\cr
NEW&Various class fusions were added.\cr
C&The {\tt galomorphisms} components which had been contained in only a few
tables were removed.\cr
***&The {\tt tomfusion} values of $L_2(25)$ and $2^5:S_6$ were corrected.\cr
***&Element orders and power maps in the table with identifier
{\tt s61p} were corrected.\cr
***&The table with identifier {\tt 2.cenc1} was removed because it was
inconsistent.\cr
% no power maps possible ...
C&Two instances of the table of $(A_6 \times A_6):2^2$ were unified.\cr
C&The tables with identifiers {\tt J2.2M4}, {\tt 2\^{}(2+4):(3x3):2\^{}2},
and {\tt 2\^{}(2+4):(S3xS3)} were unified;
the identifiers {\tt J2.2M5} and {\tt 2\^{}(2+4):(S3xS3)} can be used to
access the table.\cr
NEW&The ordinary tables of all maximal subgroups
(and their class fusions)
are now available for the groups $S_6$, $J_2.2$, $McL.2$, $Suz.2$,
$3.Suz$, $3.Suz.2$, $Sz(32)$.\cr}

\medbreak

The following changes are assigned to specific simple groups,
and shown in alphabetical order.

\medbreak

\halign{\hfil{\bf #}&\quad$#$\hfil&\quad#\hfil$\;\>$&
 \vtop{\parindent=0pt\hsize=38em\strut#\strut}\cr
NEW&A_6&:&The table of $12.A_6.2_3$ is now available.\cr
***&Fi_{22}&:&The name of the table of the $7$-th maximal subgroup
of $Fi_{22}$ was corrected from {\tt (2x2\^{}(1+8):U4(2)):2} to
{\tt (2x2\^{}(1+8)):U4(2):2};
analogously, {\tt (2x2\^{}(1+8):U4(2):2):2} was corrected to
{\tt (2x2\^{}(1+8)):(U4(2):2x2)}.\cr
C&U_3(5)&:&The table with identifier {\tt U3(5).S3} was removed;
it is replaced by the table with identifier {\tt U3(5).3.2}
whose cosets of the outer automorphism group are ordered as in the
Atlas of Finite Groups.
The identifier {\tt U3(5).S3} is now admissible for the table with
identifier {\tt U3(5).3.2}.\cr
***&U_4(3)&:&The table with identifier {\tt u4q3c} was removed;
characters and power maps of this table were erroneous.
Apparently the table was thought to be that of $3_2.U_4(3).2_3^{\prime}$,
which can be accessed with the name {\tt 3\_2.U4(3).2\_3'}.\cr
NEW&U_4(3)&:&The tables of $3_2.U_4(3).(2^2)_{133}$ and $U_4(3).(2^2)_{133}$
are now available.\cr}

\bigbreak

Last update February 02nd, 2000.

\bye

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%
%E

